---
title: "03_tidy"
output: html_document
---

## Let's start out with two data frames: x and y

```{r dfs1}
x <- data.frame(id=c(1,2,3), x=c("x1", "x2", "x3"))

x
```

```{r dfs2}
y <- data.frame(id=c(1,2,4), y=c("y1", "y2", "y4"))

y
```

## Two data frames

## left_join()


```{r two_dataframes}
library(dplyr)

left_join(x, y)
```

## Two data frames: x and y but with different column names

```{r dfs_again}
x <- data.frame(id=c(1,2,3), x=c("x1", "x2", "x3"))

x
```


```{r dfs_again2}
#| echo: TRUE
y <- data.frame(new_id=c(1,2,4), y=c("y1", "y2", "y4"))

y
```

```{r left2_join}
left_join(x, y, by=c("id"="new_id"))
```

## Watch out for repeated data

```{r left3}
x <- data.frame(id=c(1,2,3), 
                x=c("x1", "x2", "x3"))

x
```

```{r left4}
y <- data.frame(id=c(1,2,4,2), 
                y=c("y1", "y2", "y4", "y5"))

y
```

```{r}
#| echo: TRUE
left_join(x, y)
```


## stringr functions

* `str_to_upper()` `str_to_lower()` `str_to_title()`
* `str_trim()` `str_squish()`
* `str_c()`
* `str_detect()`
* `str_subset()`
* `str_sub()`

```{r str_to}
library(stringr)

test_text <- "tHiS iS A rANsOM noTE!"
```

```{r}
#| echo: TRUE
str_to_upper(test_text)
```

```{r}
#| echo: TRUE
str_to_lower(test_text)
```

```{r}
#| echo: TRUE
str_to_title(test_text)
```


## Trimming strings

```{r trim}
test_text <- "  trim both   "

test_text 
```

```{r}
#| echo: TRUE
str_trim(test_text, side="both")
```


```{r}
#| echo: TRUE
str_trim(test_text, side="left")
```


```{r}
#| echo: TRUE
str_trim(test_text, side="right")
```


```{r}
#| echo: TRUE
messy_text <- "  sometimes  you get   this "
```


```{r}
#| echo: TRUE
str_squish(messy_text)
```

## Combining strings

```{r str_c}
text_a <- "one"

text_b <- "two"

text_a

text_b
```

```{r}
#| echo: TRUE
str_c(text_a, text_b)
```

```{r}
str_c(text_a, text_b, sep="-")
```

```{r}
str_c(text_a, text_b, sep=" and a ")
```


```{r}
str_c(text_a, " and a ", text_b)
```

## Extracting strings

```{r extract}
test_text <- "Hello world"

test_text 
```


```{r}
str_sub(test_text, start = 6)
```


```{r}
str_sub(test_text, end = 5) <- "Howdy"

test_text
```

```{r}
cn <- "Kemp County, Georgia"

cn 

str_replace(cn, " County, .*", "")
```


## parse_number()

(from the readr package)

```{r parse}
library(readr)
messy_numbers <- c("$5.00", "9,343,200", "6.0%")

messy_numbers
```

```{r}
parse_number(messy_numbers)
```

# Your turn

Run this to go through the exercises

```{r, eval=F}
learnr::run_tutorial("3_a_joining", "chjr")
```

Get as far as you can in the time we have! 

2 rows x 3 columns

```{r dfs2}
df <- data.frame(id=c(1,2), x=c("a", "b"),
                 y=c("c", "d"), z=c("e", "f"))

df
```

## pivot_longer()

```{r left9}
library(tidyr)

df %>% 
  pivot_longer(cols=x:z,
               names_to="key",
               values_to="val")
```


```{r left20}
df %>% 
  pivot_longer(cols=2:4,
               names_to="key",
               values_to="val")
```


```{r left30}
df <- data.frame(state=c("TX", "NY", "FL"),
                 ducks=c(23, 39, 47),
                 fish=c(6,30,20),
                 birds=c(99,3,64))
```

## pivot_longer() again

```{r}
df
```

```{r}
df %>% 
  pivot_longer(cols=ducks:birds,
               names_to="animals",
               values_to="total")
```

## pivot_longer() again

```{r}
df
```

```{r}
#| echo: TRUE
df %>% 
  pivot_longer(cols=2:4,
               names_to="animals",
               values_to="totals")
```


```{r math1}
df <- data.frame(state=c("TX", "NY", "FL"),
                 ducks=c(23, 39, 47),
                 fish=c(6,30,20),
                 birds=c(99,3,64))
```

```{r}
df
```

```{r}
#| echo: TRUE
df %>% 
  pivot_longer(cols=ducks:birds,
               names_to="animals",
               values_to="total") %>% 
  group_by(state) %>% 
  mutate(percent=
           round(total/sum(total)*100,1))
```


## pivot_wider()

```{r wider}
df_long <- df %>% 
  pivot_longer(cols=ducks:birds,
               names_to="animals",
               values_to="total") %>% 
  group_by(state) %>% 
  mutate(percent=
           round(total/sum(total)*100,1))
```

```{r}
df_long
```

## pivot_wider()

```{r}
df_long %>% 
  pivot_wider(names_from="animals", 
              values_from="percent")
```

```{r}
df_long %>% 
  select(-total) %>% 
  pivot_wider(names_from="animals", 
              values_from="percent") %>% 
  mutate(birds_fish_diff=
           birds-fish)
```

## pivot_wider() more columns

```{r widera}
df_long <- df %>% 
  pivot_longer(cols=ducks:birds,
               names_to="animals",
               values_to="total") %>% 
  group_by(state) %>% 
  mutate(percent=
           round(total/sum(total)*100,1))
```

```{r}

df_long
```

```{r}
df_long %>% 
  pivot_wider(names_from="animals", 
              values_from=c("total", "percent")) 

```


## Lubridate for dates 



```{r dates2}
library(lubridate)

df <- data.frame(First=c("Charlie", "Lucy", "Peppermint"),
                   Last=c("Brown", "van Pelt", "Patty"),
                   birthday=c("10-31-06", "2/4/2007", "June 1, 2005"))

df
```

```{r}
df %>% 
  mutate(birthday_clean=mdy(birthday))
```


## Reading dates

| **Order of elements in date-time**     | **Parse function** |
|----------------------------------------|----------------|
| year, month, day                       | `ymd()`          |
| year, day, month                       | `ydm()`          |
| month, day, year                       | `mdy()`          |
| day, month, year                       | `dmy()`          |
| hour, minute                           | `hm()`           |
| hour, minute, second                   | `hms()`          |
| year, month, day, hour, minute, second | `ymd_hms()`      |

## Accessing date parts

| **Date component** | **Function**  |
|----------------|-----------|
| Year           | `year()`    |
| Month          | `month()`   |
| Week           | `week()`    |
| Day of year    | `yday()`    |
| Day of month   | `mday()`    |
| Day of week    | `wday()`    |
| Hour           | `hour()`    |
| Minute         | `minute()`  |
| Second         | `ymd_hms()` |
| Time zone      | `ymd_hms()` |

## Lubridate in action

```{r dates3}
df
```

```{r}
df %>% 
  mutate(birthday_clean=mdy(birthday)) %>% 
  mutate(month=month(birthday_clean)) %>% 
  mutate(year=year(birthday_clean)) %>% 
  mutate(week=week(birthday_clean))
```


# Your turn

Run this to go through the exercises

```{r, eval=F}
learnr::run_tutorial("3_b_math_pivots", "chjr")
```

Get as far as you can!



